Rapid-response Bayesian sky localization for electromagnetic follow-up of gravitational-wave candidates

### 2013-02-08 ms.tex
Adjust paths in Makefile and ms.tex after GitHub move
Merged two duplicate derivations of CRB

\documentclass{iopart}
\usepackage{iopams}
\usepackage{aas_macros}
\usepackage[colorlinks]{hyperref}
\input{acronyms}

@@

\cite{fairhurst:2009} estimated the uncertainty in the matched-filter measurement of the \ac{TOA} of a \ac{GW} signal by computing the scalar Fisher information.  We extend this result by computing the Fisher information matrix describing the uncertainties in the joint estimation of the signal amplitude and \ac{TOA}.%
%
\marginpar{This might be too much mathematical detail.  All we need to do is show that the $\rho_A\rho_A$ and $\rho_B\rho_B$ terms are unity and that the off-diagonal terms vanish.  If possible, we should cite and use Fairhurst's expression for $\mathcal{I}_{t_0,t_0}$ without re-deriving it.}

When a \ac{CBC} occurs, a detector's strain observation may be described by a linear combination of two orthogonal \ac{GW} quadratures, $a(t)$ and $b(t)$, and additive, zero\nobreakdashes-mean, \acl{WSS}, Gaussian measurement noise $n(t)$, with power spectral density $S(f)$.  Introducing the unknown, real, quadrature amplitudes, $A$ and $B$, and the unknown \ac{TOA}, $t_0$, we can write
%
\begin{equation}\label{eq:xoft}
	x(t) = A a(t - t_0) + B b(t - t_0) + n(t).
\end{equation}
%
Transformation of \Eref{eq:xoft} to the frequency domain converts the time delay by $t_0$ to a complex phase,
%
\begin{equation}\label{eq:xoff}
	\tilde{x}(f) = \left(A \tilde{a}(f) + B \tilde{b}(f)\right)e^{-2 \pi i f t_0} + \tilde{n}(f).
\end{equation}
%
The Fisher information matrix for a measurement $\mathbf{x}$ described by the unknown parameters $\boldsymbol{\theta}$ may be written as
%
\begin{equation}
	\mathcal{I}_{jk} = \mathrm{E} \, \left[
		-\frac{\partial^2 \log
			\mathcal{L}(\mathbf{x}; \boldsymbol{\theta})}
			{\partial \theta_j \theta_k}
	\right].
\end{equation}
%
First, we write the likelihood of the data $\tilde{x}(f)$ given the parameters $(A, B, t_0$):
%
\begin{equation}\label{eq:trigger-likelihood}
	\fl\mathcal{L}(x; A, B, t_0) = p(x | A, B, t_0)
		\propto \exp \left[
		- \frac{1}{2} \int_{-\infty}^\infty \frac{\left|\tilde{x}(f)
			- \left(A \tilde{a}(f)
			- B \tilde{b}(f)\right) e^{-2 \pi i f t_0}\right|^2}{S(f)} \, \mathrm{d}f
	\right].
\end{equation}
%
The normalization has been dropped because the Fisher matrix elements depend only on derivatives of the log\nobreakdashes-likelihood function and any factor that is independent of the parameters makes no contribution.%
%
\marginpar{Consider using discrete time to avoid the fiction of an infinite-dimensional multivariate Gaussian process.}
%
From hereon, unless otherwise specified the limits of integration are always $(-\infty, \infty)$ and will be dropped.  Expanding the integrand, we can write the log\nobreakdashes-likelihood function up to an additive constant as
%
\begin{eqnarray}\label{eq:trigger-likelihood}
	\log \mathcal{L}(x; A, B, t_0) \sim
	&- \frac{1}{2} \int \frac{|\tilde{x}(f)|^2}{S(f)} \mathrm{d}f \\
	&+ A \, \mathrm{Re} \int \frac{\tilde{a}^*(f) \tilde{x}(f) e^{2 \pi i f t_0}}{S(f)} \mathrm{d}f \nonumber\\
	&+ B \, \mathrm{Re} \int \frac{\tilde{b}^*(f) \tilde{x}(f) e^{2 \pi i f t_0}}{S(f)} \mathrm{d}f\nonumber\\
	&+ A B \, \mathrm{Re} \int \frac{\tilde{a}^*(f) \tilde{b}(f)}{S(f)} \mathrm{d}f \nonumber\\
	&- \frac{1}{2} A^2 \int \frac{|\tilde{a}(f)|^2}{S(f)} \mathrm{d}f
	- \frac{1}{2} B^2 \int \frac{|\tilde{b}(f)|^2}{S(f)} \mathrm{d}f.
\end{eqnarray}
%
We assume that $\tilde{a}(f)$ and $\tilde{b}(f)$ are orthogonal and have the same norm,
%
\begin{equation}
	\int \frac{\tilde{a}^*(f) \tilde{b}(f)}{S(f)} \mathrm{d}f = 0,
	\int \frac{|\tilde{a}(f)|^2}{S(f)} \mathrm{d}f =
	\int \frac{|\tilde{b}(f)|^2}{S(f)} \mathrm{d}f =
	\sigma^2.
\end{equation}
%
After we apply these substitutions, and drop the $|\tilde{x}(f)|^2$ term that is independent of $A$, $B$, or $t_0$, we can write the log\nobreakdashes-likelihood function as
%
\begin{eqnarray}\label{eq:trigger-likelihood-simplified}
	\log \mathcal{L}(x; A, B, t_0) \sim
	& A \, \mathrm{Re} \int \frac{\tilde{a}^*(f) \tilde{x}(f) e^{2 \pi i f t_0}}{S(f)} \mathrm{d}f 
	&+ B \, \mathrm{Re} \int \frac{\tilde{b}^*(f) \tilde{x}(f) e^{2 \pi i f t_0} }{S(f)} \mathrm{d}f\nonumber\\
	&- \frac{1}{2} A^2 \sigma^2 - \frac{1}{2} B^2 \sigma^2.
\end{eqnarray}
%
The necessary partial derivatives are
%
\begin{eqnarray*}
	\frac{\partial^2 \log \mathcal{L}}{\partial A \partial B} &= 0 \\
	\frac{\partial^2 \log \mathcal{L}}{\partial A \partial t_0} &=
	2 \pi i \int \frac{\left(\tilde{a}^*(f) \tilde{x}(f) e^{2 \pi i f t_0} - \tilde{a}(f) \tilde{x}^*(f) e^{-2 \pi i f t_0}\right)}{S(f)} f \mathrm{d}f \\
	\frac{\partial^2 \log \mathcal{L}}{\partial B \partial t_0} &=
	2 \pi i \int \frac{\left(\tilde{b}^*(f) \tilde{x}(f) e^{2 \pi i f t_0} - \tilde{b}(f) \tilde{x}^*(f) e^{-2 \pi i f t_0}\right)}{S(f)} f \mathrm{d}f \\
	\frac{\partial^2 \log \mathcal{L}}{\partial A^2} &=
	\frac{\partial^2 \log \mathcal{L}}{\partial B^2} = - \sigma^2 \\
	\frac{\partial^2 \log \mathcal{L}}{\partial {t_0}^2} &=
	2 \pi i A \int \frac{\left(\tilde{a}^*(f) \tilde{x}(f) e^{2 \pi i f t_0} - \tilde{a}(f) \tilde{x}^*(f) e^{-2 \pi i f t_0}\right)}{S(f)} f \mathrm{d}f \\
	&+ 2 \pi i B \int \frac{\left(\tilde{b}^*(f) \tilde{x}(f) e^{2 \pi i f t_0} - \tilde{b}(f) \tilde{x}^*(f) e^{-2 \pi i f t_0}\right)}{S(f)} f \mathrm{d}f \\
	&- (2 \pi)^2 A \int \frac{\left(\tilde{a}^*(f) \tilde{x}(f) e^{2 \pi i f t_0} + \tilde{a}(f) \tilde{x}^*(f) e^{-2 \pi i f t_0}\right)}{S(f)} f^2 \mathrm{d}f \\
	&- (2 \pi)^2 B \int \frac{\left(\tilde{b}^*(f) \tilde{x}(f) e^{2 \pi i f t_0} + \tilde{b}(f) \tilde{x}^*(f) e^{-2 \pi i f t_0}\right)}{S(f)} f^2 \mathrm{d}f.
\end{eqnarray*}
%
Finally, taking the expectation values of the partial derivatives, we find
%
\begin{eqnarray*}
	\mathcal{I}_{A,B} &= \mathcal{I}_{A,t_0} = \mathcal{I}_{B,t_0} = 0 \\
	\mathcal{I}_{A,A} &= \mathcal{I}_{B,B} = \sigma^2 \\
	\mathcal{I}_{t_0,t_0} &=
	- 2 \pi i A^2 \int \frac{|\tilde{a}(f)|^2}{S(f)} f\mathrm{d}f
	- 2 \pi i B^2 \int \frac{|\tilde{B}(f)|^2}{S(f)} f\mathrm{d}f \\
	&+ (2 \pi)^2 \left(A^2 \int \frac{|\tilde{a}(f)|^2}{S(f)} f^2\mathrm{d}f +
	B^2 \int \frac{|\tilde{b}(f)|^2}{S(f)} f\mathrm{d}f \right) \\
	&+ (2\pi)^2 2 A B \, \mathrm{Re} \int \frac{\tilde{a}^*(f)\tilde{b}(f)}{S(f)} f^2\mathrm{d}f.
\end{eqnarray*}
%
$\mathcal{I}_{A,t_0}$, $\mathcal{I}_{B,t_0}$, and the first two terms of $\mathcal{I}_{t_0,t_0}$ vanish because the integrands $|\tilde{a}(f)|^2 f / S(f)$ and $|\tilde{b}(f)|^2 f / S(f)$ are odd about $f = 0$.  We assume that $\tilde{a}(f)$ and $\tilde{b}(f)$ are orthogonal with respect to $\emph{any}$ real weight, including both $S(f)$ and $S(f) / f^2$, so the last term in $\mathcal{I}_{t_0,t_0}$ vanishes as well.  Further,%
%
\marginpar{This is an awkward place to make these substitutions.  Also, they are poorly motivated.  Here's what this substitution actually does: the matched\nobreakdashes-filter output has an expected value of $A \sigma = \rho_A$.}
%
we make the substituions
%
\begin{eqnarray*}
	\rho_A &= A \sigma \\
	\rho_B &= B \sigma \\
	f_\mathrm{rms} &= \left[ \int \frac{|\tilde{a}(f)|^2}{S(f)} f^2\mathrm{d}f \right] ^ {1/2} \left[\int \frac{|\tilde{a}(f)|^2}{S(f)} \mathrm{d}f \right] ^ {-1/2} \\
	&=  \left[ \int \frac{|\tilde{b}(f)|^2}{S(f)} f^2\mathrm{d}f \right] ^ {1/2} \left[\int \frac{|\tilde{b}(f)|^2}{S(f)} \mathrm{d}f \right] ^ {-1/2}.
\end{eqnarray*}
%
Then, we have
%
\begin{eqnarray*}
	\mathcal{I}_{\rho_A,\rho_B} &= \mathcal{I}_{\rho_A,t_0} = \mathcal{I}_{\rho_B,t_0} = 0 \\
	\mathcal{I}_{\rho_A,\rho_A} &= \mathcal{I}_{\rho_B,\rho_B} = 1 \\
	\mathcal{I}_{t_0,t_0} &= (2 \pi f_\mathrm{rms})^2 ({\rho_A}^2 + {\rho_B}^2).
\end{eqnarray*}
%
The Cram\'{e}r-Rao inequality says that the inverse of the Fisher information matrix is a lower bound on the covariance of any unbiased estimator.  The Fisher matrix for this estimation problem has been shown to be diagonal, so we can write
%
\begin{eqnarray}\label{eq:cramer-rao}
	\mathrm{var}\, (\hat{\rho}_B) &\geq {\mathcal{I}_{\rho_A,\rho_A}}^{-1} = 1 \\
	\mathrm{var}\, (\hat{\rho}_B) &\geq {\mathcal{I}_{\rho_B,\rho_B}}^{-1} = 1 \\
	\mathrm{var}\, (\hat{t}_0) &\geq {\mathcal{I}_{t_0,t_0}}^{-1} = (2 \pi f_\mathrm{rms} \sqrt{{\rho_A}^2 + {\rho_B}^2})^{-2}.
\end{eqnarray}

@@

\section{Equations that need to be organized}

\begin{equation}
	u = \cos \iota
\end{equation}

\begin{equation}
	\mathbf{G} =
	\left(
		\begin{array}{cccc}
			F_+^1 D_1^1 & F_+^2 D_1^2 & \cdots & F_+^N D_1^N \\
			F_\times^1 D_1^1 & F_\times^2 D_1^2 & \cdots & F_\times^N D_1^N
		\end{array}
	\right)
\end{equation}

\begin{equation}
	\mathbf{H} = \mathbf{G}\mathbf{G}^\mathsf{T} =
	\left(
		\begin{array}{cc}
			\sum \left(F_+^i D_1^i\right)^2 & \sum \left(F_+^i D_1^i\right)\left(F_\times^i D_1^i\right) \\
			\sum \left(F_+^i D_1^i\right)\left(F_\times^i D_1^i\right) & \sum \left(F_\times^i D_1^i\right)^2
		\end{array}
	\right)
\end{equation}

\begin{equation}
	\mathbf{R} =
	\left(
		\begin{array}{cc}
			\cos \phi & \sin \phi \\
			-\sin \phi & \cos \phi
		\end{array}
	\right)
	\left(
		\begin{array}{cc}
			\frac{1}{2} (1 + u^2) & 0 \\
			0 & u
		\end{array}
	\right)
	\left(
		\begin{array}{cc}
			\cos \psi & \sin \psi \\
			-\sin \psi & \cos \psi
		\end{array}
	\right) \mathbf{G}
\end{equation}

\begin{equation}
	p = \Tr \mathbf{R}^\mathsf{T} \mathbf{R}
	= \frac{1}{8} \left\{
		\left(u^4 + 6 u^2 + 1\right) \left(H_{11} + H_{22}\right) + (1 - u^2)^2
		\left[ \left( H_{11} - H_{22} \right) \cos 2\psi + 2 H_{12} \sin 2\psi \right]
	\right\}
\end{equation}

\begin{equation}\fl
	q = \sqrt{(ab - cd) (u + u^3) + \frac{1}{8} \left\{ (a^2 + b^2 + c^2 + d^2)(u^4 + 6u^2 + 1) + (1 - u^2)^2 \left[ (a^2 - b^2 - c^2 + d^2) \cos 2\psi + 2(ac + bd) \sin 2\psi \right] \right\}}
\end{equation}

@@

\section*{References}
\bibliographystyle{iopart-num}
\bibliography{apj-jour,telescope}

>>>>

\documentclass{iopart}
\usepackage{iopams}
\usepackage{bib/aas_macros}
\usepackage[colorlinks]{hyperref}
\input{ligo-acronyms/acronyms}

@@

\cite{fairhurst:2009} estimated the uncertainty in the matched-filter measurement of the \ac{TOA} of a \ac{GW} signal by computing the scalar Fisher information.  We extend this result by computing the Fisher information matrix describing the uncertainties in the joint estimation of the signal amplitude and \ac{TOA}.

When a \ac{CBC} occurs, a detector's strain observation may be described by a linear combination of two orthogonal \ac{GW} quadratures and additive, zero\nobreakdashes-mean, \acl{WSS}, Gaussian measurement noise $n(t)$, with one-sided power spectral density $S(f)$. In the frequency domain, the detector's observation $\tilde x (\omega)$ is given by
%
\begin{equation}\label{eq:signal-model}
	\tilde x (\omega) = \tilde h (\omega, \rho, \gamma, \tau) + \tilde n (\omega) = \rho e^{\imath (\gamma - \omega \tau)} \tilde u (\omega) + \tilde n (\omega).
\end{equation}
%
where $\tilde h (\omega)$ is the signal content, $\tilde n (\omega)$ is the detector's noise, $\tilde u (\omega)$ is the frequency-domain post-Newtonian inspiral signal, $\rho$ is the signal's amplitude, $\gamma$ is a phase angle the encodes the relative strength of the two \ac{GW} quadratures, and $\tau$ is the time of arrival relative to an arbitrary time base.

The Fisher information matrix for a measurement $\tilde x$ described by the unknown parameter vector $\boldsymbol{\theta}$ may be written as
%
\begin{equation}\label{eq:general-fisher-matrix}
	\mathcal{I}_{jk} = \mathrm{E} \, \left[
		-\frac{\partial^2 \log
			\mathcal{L}(\tilde x ; \boldsymbol\theta)}
			{\partial \theta_j \theta_k}
	\right].
\end{equation}
%
This equation involves double derivatives, but substantial simplification is possible when the likelihood is Gaussian. First, we write the likelihood of the data $\tilde x (\omega)$ given the parameter vector $\boldsymbol\theta = (\rho, \gamma, \tau)$:
%
\begin{equation}\label{eq:gaussian-likelihood}
	\mathcal{L}(\tilde x; \boldsymbol\theta) = p(\tilde x | \boldsymbol\theta)
		\propto \exp \left[
		- \frac{1}{2} \int_0^\infty \frac{\left|\tilde x (\omega)
			- \tilde h (\omega; \boldsymbol\theta) \right|^2}{S(\omega)} \, \mathrm{d}\omega
	\right].
\end{equation}
%
The normalization has been dropped because the Fisher matrix elements depend only on derivatives of the log\nobreakdashes-likelihood function and any factor that is independent of the parameters makes no contribution. Now, substituting \eref{eq:gaussian-likelihood} into \eref{eq:general-fisher-matrix} gives us
%
\begin{equation}\label{eq:gaussian-fisher-matrix}
	\mathcal{I}_{jk} = \int_0^\infty \Re \left[
        \left(\frac{\partial \tilde h}{\partial \theta_j}\right)^*
        \left(\frac{\partial \tilde h}{\partial \theta_k}\right)
	\right] \frac{1}{S(\omega)} \, \mathrm{d}\omega.
\end{equation}
%
where $\Re$ denotes the real part and $^*$ the complex conjugate. This form is more useful because it involves only first derivatives, and of the signal $\tilde h (\omega)$ rather than the entire observation $\tilde x (\omega)$.

With \eref{eq:gaussian-fisher-matrix} in hand, we return to our particular signal model, \Eref{eq:signal-model}. The necessary partial derivatives are
%
\begin{eqnarray*}
    \frac{\partial \tilde h}{\partial \rho} &= e^{\imath (\gamma - \omega \tau)} \tilde u (\omega) \\
    \frac{\partial \tilde h}{\partial \gamma} &= \imath \, \rho \, e^{\imath (\gamma - \omega \tau)} \tilde u (\omega) \\
    \frac{\partial \tilde h}{\partial \tau} &= -\imath \, \omega \, \rho \, e^{\imath (\gamma - \omega \tau)} \tilde u (\omega).
\end{eqnarray*}
%
The Fisher matrix elements are
%
\begin{eqnarray*}
    \mathcal{I}_{\rho\rho} &= \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \, \mathrm{d}\omega \\
	\mathcal{I}_{\gamma\gamma} &= \rho^2 \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \, \mathrm{d}\omega \\
	\mathcal{I}_{\tau\tau} &= \rho^2 \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \omega^2 \, \mathrm{d}\omega \\
	\mathcal{I}_{\gamma\tau} &= -\rho^2 \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \omega \, \mathrm{d}\omega = \mathcal{I}_{\tau\gamma} \\
	\mathcal{I}_{\rho\gamma} &= \mathcal{I}_{\gamma\rho} = \mathcal{I}_{\rho\tau} = \mathcal{I}_{\tau\rho} = 0.
\end{eqnarray*}
%
If the waveform $\tilde u (\omega)$ is normalized such that $\mathcal{I}_{\rho\rho} = 1$, we can introduce the \textit{\ac{SNR}-weighted moments of angular frequency},
%
\begin{equation}\label{eq:angular-frequency-moments}
    \overline{\omega^k} =
        \left[ \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \omega^k \, \mathrm{d}\omega \right]
        \left[ \int_0^\infty \frac{|\tilde u (\omega)|^2}{S(\omega)} \, \mathrm{d}\omega \right]
\end{equation}
%
Then, we have
%
\begin{eqnarray*}
	\mathcal{I} = \bordermatrix{
        ~ & \rho & \gamma & \tau \cr
        \rho & 1 & 0 & 0 \cr
        \gamma & 0 & \rho^2 & -\rho^2 \overline{\omega} \cr
        \tau & 0 & -\rho^2 \overline{\omega} & \rho^2 \overline{\omega^2}
    }.
\end{eqnarray*}

@@

Now we will perform four changes of variables. The first change of variables is from $C$ to $\rho = \sqrt{M_0}C$. This change is intended to make the variance of the absolute value of the amplitude equal to one. 
\[
\mathcal{I} =
\left( {\begin{array}{ccc}
 1 & 0 & 0 \\
 0 & \rho^2 & -2\pi\rho^2 N_1 \\
 0 & -2\pi\rho^2 N_1 & 4\pi^2 \rho^2 N_2
 \end{array} } \right)
\]

The second from $t$ to $\phi_t = t2\pi \sqrt{N_2}$ is meant to get rid of $t_0$,
\[
\mathcal{I} =
\left( {\begin{array}{ccc}
 1 & 0 & 0 \\
 0 & \rho^2 & -\rho^2\frac{N_1}{\sqrt{N_2}} \\
 0 & -\rho^2\frac{N_1}{\sqrt{N_2}} & \rho^2
 \end{array} } \right)
\]

The third introduces two new variables, $\phi_s = \frac{1}{\sqrt{2}}(\phi + \phi_t)$ and $\phi_d = \frac{1}{\sqrt{2}}(\phi - \phi_t)$. This change of variables diagonalizes the matrix.
\[
\mathcal{I} =
\left( {\begin{array}{ccc}
 1 & 0 & 0 \\
 0 & (1 - \frac{N_1}{\sqrt{N_2}})\rho^2 & 0 \\
 0 & 0 & (1 + \frac{N_1}{\sqrt{N_2}})\rho^2
 \end{array} } \right)
\]

Now we'll do a final change of variables to make the phase parameters have the same variance. We let $\alpha_s = \phi_s \sqrt{1 - \frac{N_1}{\sqrt{N_2}}}$ and $\alpha_d = \phi_d \sqrt{1 - \frac{N_1}{\sqrt{N_2}}}$ which gives us the matrix
\[
\mathcal{I} =
\left( {\begin{array}{ccc}
 1 & 0 & 0 \\
 0 & \rho^2 & 0 \\
 0 & 0 & \rho^2
 \end{array} } \right)
\]

@@

\section*{References}
\bibliographystyle{iopart-num}
\bibliography{apj-jour,bib/telescope}
